Efficient Calculation of Polynomial and Interaction Features on Sparse Matrices

### 2015-12-03 paper/paper2.tex

\section{Approach}
Let the list of nonzero columns for a given row $\vec{x}$ be denoted by $N_{zc}$. The nonzero second degree 
interaction features are simply the products of all combinations of two elements whose 
columns are in $N_{zc}$. However, to properly place an interaction feature into the correct column, a mapping from the column 
index pairs of $N_{zc}$ into the columns of the interaction matrix is needed. The mapping is 
from pairs $(a, b)$,  where $a$ and $b$ are in $1,2,..., D$, and $a < b$, to $1,2,..., \frac{D^2-D}{2}$,  Such a mapping essentially consists of mapping the indices of entries in the upper triangle of a matrix to indices in a flat 
list. We now describe the construction of such a mapping.

\subsection{Mapping construction}
\textbf{TODO: convert all $i$ and $j$ indices into $a$ and $b$.--jfh}

We seek a map from matrix indices $(i, j)$ (with $i < j$ and $0 \le i < D$) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by 
\begin{align}
\begin{bmatrix}
x & 0 & 1 & 3 \\
x & x & 2 & 4 \\
x & x & x & 5 \\
x & x & x & x
\end{bmatrix}
\label{eq:4x4mat}
\end{align}
where the entry at location $(i,j)$ displays the value $f(i, j)$.

To simplify slightly, we introduce a notation for the $n$th triangular number, 
\begin{align}
T_2(n) = \frac{n(n+1)}{2}
\end{align}
\noindent
The subscript $2$ is there to indicate that these are triangles in two dimensions; we'll use $T_3(n)$ to indicate the $n$th tetrahedral number, and so on for higher dimensions.

Observe that in Equation~\ref{eq:4x4mat}, each entry in column $j$ lies in the range
\begin{align}
T_2(j-1) &\le e < T_2(j).
\end{align}
\noindent
And in fact, the entry in the $i$th row of that column is just $i + T_2(j-1)$. Thus we have
\begin{align}
f(i, j) 
&= i + T_2(j-1)\\
&= i + \frac{(j-1)j}{2}\\
&=  \frac{2i + j^2-j}{2}.
\end{align}

@@

\subsubsection{Other indices}
With one-based indexing in both the domain and codomain, the formula above becomes
\begin{align}
f_1(i, j) &= 1+ f(i-1, j-1) \\
&= 1+ f(i-1, j-1) \\
& = \frac{2 + 2(i-1) + (j-1)^2-(j-1)}{2}\\
& = \frac{2i + j^2 - 3j + 2}{2}
\end{align}
\subsubsection{Higher dimensions}
To handle three-way interactions, we need to map triples of indices in a 3-index array to a flat list, and similarly for higher-order interactions.

For three indices, $i,j,k$, with $i < j < k$ and $0 \le i,j,k < D$, we have a similar recurrence. Calling the mapping $g$, we have 
\begin{align}
g(i,j,k) = i + T_2(j-1) + T_3(k-1);
\end{align}
if we define $T_1(i) = i$, then this has the very regular form
\begin{align}
g(i,j,k) = 1 + T_1(i-1) + T_2(j-1) + T_3(k-1);
\end{align}
and from this, the generalization to higher dimensions is straightforward. The formulas for ``higher triangular numbers'', i.e., those defined by
\begin{align}
T_k(n) &= \sum_{i=1}^n T_{k-1}(n)
\end{align}
for $k > 1$ can be determined inductively. For $k = 3$, the result is 
\begin{align}
T_3(n) &= \sum_{i=1}^n T_{2}(n)\\
&= \frac{n^3 + 3n^2 + 2n}{6},
\end{align}
so that the formula for 3-way interactions, with zero-based indexing, becomes 
\begin{align}
g(i, j, k) &= 1 + (i-1) + \frac{(j-1)j}{2} + \\
& \frac{(k-1)^3 + 3(k-1)^2 + 2(k-1)}{6}. 
\end{align}
\subsection{Inversion}
We not only want to be able to compute the ``flat index'' $f(i,j)$ corresponding to a particular index pair $(i, j)$, we'd like, given a value $N = f(i, j)$, to be able to determine $i$ and $j$. We know, given $N$, that $f(0, j) \le N \le f(j,j-1)$, i.e., that 
\begin{align}
\frac{j^2 - 3j + 2}{2} &\le N &\le \frac{2(j-1) + j^2 - 3j + 2}{2}\\
{j^2 - 3j + 2} &\le 2N &\le {j^2 - j}.
\end{align}
If we find the (real number) value $j_{*}$ for which $j_{*}^2 - j_{*} = 2N$, then we know that $j = \lfloor j_{*}\rfloor$. Finding $j_{*}$ amounts to solving a quadratic, whence
\begin{align}
j_{*} = \frac{1 \pm \sqrt{1 + 8N}}{2}.
\end{align}
Clearly the larger value, 
\begin{align}
j_{*} = \frac{1 + \sqrt{1 + 8N}}{2},
\end{align}
is the one we seek; we set $j = \lfloor j_{*}\rfloor$, and compute $i = N - \frac{j^2 - j}{2}$.

In practice, repeatedly computing square roots may be annoying; instead, when we compute $N = f(i,j)$, we can record the values of $i$ and $j$ along with the interaction features, making inversion a simple matter of looking up these values.

%To construct this mapping, consider a matrix of size $D \times D$ where %cell $(a,b)$ represents the
%polynomial feature between indices $a$ and $b$ for $a,b \in I$ for vector $\vec{x}$.

%\renewcommand{\kbldelim}{(}% Left delimiter
%\renewcommand{\kbrdelim}{)}% Right delimiter
%\[
%  \kbordermatrix{
%             & x_0    & x_1   & \cdot   & \cdot   &  \cdot  &  x_{D-1} \\
%     x_0     & 0      & 1     & \cdot   &  \cdot  &  \cdot  &  D-1     \\
%     x_1     & -      & D     & \cdot   &  \cdot  &  \cdot  &  2D-2    \\
%     \cdot   & \cdot  & \cdot &  \cdot  &      &      &  \cdot       \\
%     \cdot   & \cdot  &       &  \cdot  &  \cdot  &      &  \cdot       %\\
%     \cdot   & \cdot  &       &         &  \cdot  &  \cdot  &  \cdot       %\\
%    x_{D-1}  & -      & \cdot &  \cdot  & \cdot   & -    &  \frac{D^2+D}%{2}-1\\
  %}
%\]

%Again, any bijective mapping $I \mapsto O$ would suffice, but this %particular mapping could
%be considered one of the more natural options as it maps $I$ onto $O$ %where both $I$ and $O$ are sorted in ascending order.
%As can be seen, the required mapping can be viewed as mapping the upper %triangular matrix (including 
%the diagonal), whose indicies are $I$, onto $O$. The construction of this %mapping will now
%be given.

%The mapping can be modified slightly to yield interaction features %instead of polynomial features
%by noting that the section of the $D \times D$ mapping matrix that will %yield interaction
%features differs only in that its diagonal is excluded. The mapping is %therefore gotten by
%taking the polinomial mapping and subtracting $i+1$.

>>>>

\section{Approach}
Let the list of nonzero columns for a given row $\vec{x}$ be denoted by $N_{zc}$. The nonzero second degree 
interaction features are simply the products of all combinations of two elements whose 
columns are in $N_{zc}$. However, to properly place an interaction feature into the correct column, a mapping from the column 
index pairs of $N_{zc}$ into the columns of the interaction matrix is needed. The mapping is 
from pairs $(i, j)$,  where $i$ and $j$ are in $0, 1, \ldots , D-1$, and $i < j$, to $0, 1,2,..., \frac{D^2-D}{2} - 1$,  Such a mapping essentially consists of mapping the indices of entries in the upper triangle of a matrix to indices in a flat 
list. We now describe the construction of such a mapping.

\subsection{Mapping construction}

We seek a map from matrix indices $(i, j)$ (with $i < j$ and $0 \le i < D$) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by 
\begin{align}
\begin{bmatrix}
x & 0 & 1 & 3 \\
x & x & 2 & 4 \\
x & x & x & 5 \\
x & x & x & x
\end{bmatrix}
\label{eq:4x4mat}
\end{align}
where the entry in row $i$, column $j$, displays the value $f(i, j)$.

To simplify slightly, we introduce a notation for the $n$th triangular number, 
\begin{align}
T_2(n) = \frac{n(n+1)}{2}
\end{align}
\noindent
The subscript $2$ indicates that these are triangles in two dimensions; we'll use $T_3(n)$ to indicate the $n$th tetrahedral number, and so on for higher dimensions.

Observe that in Equation~\ref{eq:4x4mat}, each entry in column $j$ (for $j > 0$) lies in the range
\begin{align}
T_2(j-1) &\le e < T_2(j).
\end{align}
\noindent
And in fact, the entry in the $i$th row of that column is just $i + T_2(j-1)$. Thus we have
\begin{align}
f(i, j) 
&= i + T_2(j-1)\\
&= i + \frac{(j-1)j}{2}\\
&=  \frac{2i + j^2-j}{2}.
\end{align}

@@

\subsubsection{Other indices}
With one-based indexing in both the domain and codomain, the formula above becomes
\begin{align}
f_1(i, j) &= 1+ f(i-1, j-1) \\
&= 1+ f(i-1, j-1) \\
& = \frac{2 + 2(i-1) + (j-1)^2-(j-1)}{2}\\
& = \frac{2i + j^2 - 3j + 2}{2}
\end{align}

\subsubsection{Polynomial Features}
For polynommial features, we seek a map from matrix indices $(i, j)$ (with $i \le j$ and $0 \le i < D$) to numbers $g(i, j)$ with $0 \le f(i, j) < \frac{D(D+1)}{2}$, one that follows the pattern indicated by 
\begin{align}
\begin{bmatrix}
 0 & 1 & 3 & 6 \\
 x & 2 & 4 & 7\\
 x & x & 5 & 8 \\
 x & x & x & 9
\end{bmatrix}
\label{eq:4x4mat-poly}
\end{align}
i.e., essentially the same task as before, except that the diagonal is included. One can regard all but the last column of entries in Equation~\ref{eq:4x4mat-poly} as corresponding to the entries in Equation~\ref{eq:4x4mat}, but shifted to the left. Thus the formula for $g(i, j)$ is simply the formula for $f$, shifted by 1, i.e., 
\begin{align}
g(i, j) &= f(i, j+1)  \\
&=  \frac{2i + (j+1)^2-(j+1)}{2}\\
&=  \frac{2i + j^2 + j + 1)}{2}.
\end{align}
Alternatively, we can write this as
\begin{align}
g(i, j) &= i + T_2(j),
\end{align}
\noindent
and get the same result.

\subsubsection{Higher dimensions}
To handle three-way interactions, we need to map triples of indices in a 3-index array to a flat list, and similarly for higher-order interactions.

For three indices, $i,j,k$, with $i < j < k$ and $0 \le i,j,k < D$, we have a similar recurrence. Calling the mapping $h$, we have 
\begin{align}
h(i,j,k) = i + T_2(j-1) + T_3(k-2);
\end{align}
if we define $T_1(i) = i$, then this has the very regular form
\begin{align}
h(i,j,k) =  T_1(i) + T_2(j-1) + T_3(k-2);
\end{align}
and from this, the generalization to higher dimensions is straightforward. The formulas for ``higher triangular numbers'', i.e., those defined by
\begin{align}
T_k(n) &= \sum_{i=1}^n T_{k-1}(n)
\end{align}
for $k > 1$ can be determined inductively. For $k = 3$, the result is 
\begin{align}
T_3(n) &= \sum_{i=1}^n T_{2}(n)\\
&= \frac{n^3 + 3n^2 + 2n}{6},
\end{align}
so that the formula for 3-way interactions, with zero-based indexing, becomes 
\begin{align}
h(i, j, k) &= 1 + (i-1) + \frac{(j-1)j}{2} + \\
& \frac{(k-2)^3 + 3(k-2)^2 + 2(k-2)}{6}. 
\end{align}
\subsubsection{Higher-dimension polynomial features}
For the case where we include the diagonal in higher dimensions, we must shift $j$ by $1$, $k$ by $2$, and so on, and the formula becomes
\begin{align}
\ell(i,j,k) =  T_1(i) + T_2(j) + T_3(k),
\end{align}
with analogous formulas for higher degree polynomial interactions.

\subsection{Inversion}
We not only want to be able to compute the ``flat index'' $f(i,j)$ corresponding to a particular index pair $(i, j)$, we'd like, given a value $N = f(i, j)$, to be able to determine $i$ and $j$. We know, given $N$, that $f(0, j) \le N \le f(j,j-1)$, i.e., that 
\begin{align}
\frac{j^2 - 3j + 2}{2} &\le N \le \frac{2(j-1) + j^2 - 3j + 2}{2}\\
{j^2 - 3j + 2} &\le 2N \le {j^2 - j}.
\end{align}
If we find the (real number) value $j_{*}$ for which $j_{*}^2 - j_{*} = 2N$, then we know that $j = \lfloor j_{*}\rfloor$. Finding $j_{*}$ amounts to solving a quadratic, whence
\begin{align}
j_{*} = \frac{1 \pm \sqrt{1 + 8N}}{2}.
\end{align}
Clearly the larger value, 
\begin{align}
j_{*} = \frac{1 + \sqrt{1 + 8N}}{2},
\end{align}
is the one we seek; we set $j = \lfloor j_{*}\rfloor$, and compute $i = N - \frac{j^2 - j}{2}$. 
In practice, repeatedly computing square roots may be annoying; instead, when we compute $N = f(i,j)$, we can record the values of $i$ and $j$ along with the interaction features, making inversion a simple matter of looking up these values.

Similarly, for polynomial features, to invert 
\begin{align}
g(i, j) &=  \frac{2i + j^2 + j + 1)}{2}
\end{align}
we know that if $N = g(i,j)$, then $g(0,j) \le N \le g(j,j)$, which tells us that 
\begin{align}
2N \le j^2 + 3j + 1, \text{ so}\\
0 \le j^2 + 3j + (1-N^2)
\end{align}. 
Solving for $j$ when this last inequality is an equality, we get 
\begin{align}
j_{*} = \frac{-3 \pm \sqrt{ 5 - N^2 }}{2}; 
\end{align}
Once again, we set $j = \lfloor j_{*}\rfloor$, and compute $i = N - \frac{j^2 + j + 1}{2}$.

Inverting the higher-order formulas, which involve cubics, quartics, etc., is infeasible, but storing the inverse map as described above may be practical, especially for sparse feature sets.
